Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Guideline for International \emph{and Domestic} Students

### 2024-02-21 faculty.tex
updates

\mainmatter
\chapter{Should You Apply to a CS PhD Program in the US?}\label{sec:should}

\epigraph{\vspace{-0.2in} Don't make fun of graduate students. They just made a terrible life choice.}{\textsc{The Simpsons}}

First, I want to emphasize that PhD students in Computer Science \emph{do not} need to worry about funding, especially at good R1
universities in the US. If you are admitted, you will almost certainly
receive \emph{full funding} to support your study, including tuition,
health insurance, and stipend (monthly salary). Moreover, depending on the university,
you may even receive additional benefits such as summer pay, laptops, (conference/workshop) traveling. \autoref{sec:funding} provides more details on funding.

Second, applying to a good US university \emph{should not} be any harder than at schools in other countries. In fact, it might even be more flexible since CS PhD in the US \emph{do not} require having an MS or a topic/proposal/adviser in advance. If you believe you have a chance in other countries, e.g., South Korea, Singapore, Germany, UK, Japan, and Australia, then you will surely have a chance in the US as well. \autoref{sec:non-us-differences} compares CS PhD study in the US to other countries.

\begin{commentbox}[Vu:]
  One of the reasons I create this document is that my colleagues at GMU are interested in recruiting Vietnamese students and are surprised when seeing very few applications in Vietnam (see \autoref{sec:ack}). Each year our CS program receives more than 350 PhD applications, most of which are international but only 3--4 are from Vietnam. In general the number of CS PhD applications from Vietnam to US universities is very low and more would be welcome.
\end{commentbox}

\begin{domesticbox}[Domestic students:] If you're a domestic student, you have several advantages in your application.  First, we already know about your school (see \autoref{sec:your-school}), and you are also more familiar with the US education system and culture.  You also have more opportunities for funding, e.g., through government scholarships for US citizens and residents. Finally, compared to international students, PhD applications in CS from domestic students are few and that can really help your case.
\end{domesticbox}

%% Vu: *what's a PhD?*  This [[https://matt.might.net/articles/phd-school-in-pictures/][series of pictures]] from [[https://matt.might.net][Matt Might]] illustrates what a PhD means.
%% \end{commentbox}

\section*{Additional Resources}
\begin{itemize}
  \item \href{https://parentheticallyspeaking.org/articles/us-cs-phd-faq/}{Getting a Computer Science PhD in the USA} by Shriram Krishnamurthi
  \item  \href{https://pg.ucsd.edu/PhD-application-tips.htm}{Blog post on PhD Admission} by Philip Guo
  \item   \href{https://talkingtorobots.com/yonatanbisk.html}{PhD Admission post} by Yonatan Bisk (customized for CMU admission)
\end{itemize}

\chapter{How is Your Application Evaluated?}\label{sec:evalapps}

\epigraph{How is education supposed to make me feel smarter? Besides, every time I learn something new, it pushes some of the old stuff out of my brain. Remember when I took that home wine making course, and I forgot how to drive?}{\textsc{The Simpsons}}

After you submit your PhD application (usually around December), it will first be screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters? Usually, this screen process is done through a central university system (i.e., not by the CS department).  If you pass this screen, your application will be forwarded to the CS department for further evaluation. If you don't, then the system will inform you what is missing and what you need to do.

\begin{commentbox}[Hakan:]
  At GMU, for full consideration, students should make sure to submit \textbf{ALL} required documents by the application deadline, and should never assume that some required documents (such as official TOEFL scores or official diplomas/transcripts) will be waived by the admissions office. If something is listed and not marked as ``optional'', it is mandatory and they should plan for submitting all those.
\end{commentbox}

Then, applications are reviewed by a \textbf{PhD admission committee} (\emph{adcom}) that consists of faculty members in CS (in some cases the committee can involve affiliated faculty from different disciplines). These faculty have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depends on the department size. At GMU, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review 25--30 applications. Note that GMU, UMass, and probably many other schools, have separate adcom for MS programs.

Each application is assigned to about \emph{three} faculty members, who will evaluate your profile and reach a consensus.  Note that while the assigned reviewers are likely the main ones deciding your application, \emph{every} faculty in the department will have access to your application and can provide inputs and opinions on your profile.

The PhD adcom typically involves assistant professors in the department (see \autoref{sec:faculty-types} for various faculty titles). This provides junior faculty the opportunities to recruit students. The chair of the committee will be a senior professor, but they likely will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty.

At GMU, we usually decide that a full-time PhD candidate is either (i) admit with funding (TA or RA, see \autoref{sec:funding}) or (ii) rejected. In other words, in most cases, we either
admit you with full funding, or we reject your application. In some rare cases, we admit
without funding because you have funding on your own, e.g.,
supported by your government or having external fellowships. We justify
our decision with a summary of your application, where we list
strengths (e.g., a well-known school) and weaknesses (e.g., weak
LORs).

% \didi{Is there more information on typical strengths and common weaknesses of applications. This is especially useful to sophomore and junior students as they still have time to work on those strengths.}
% tvn: the main thing is research experiences

\paragraph{Why we do not waive application fee?}  This is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

In my opinion, requiring applicants to pay the fee helps ensure their seriousness, as it filters out non-serious candidates. Also, if the application process were free for everyone, we would receive an overwhelming number of applications to review.

% \begin{commentbox}[Hung]
%   The application fee varies between schools, and is about \$70--\$100. I found this number pretty ridiculous. In most cases, there is no proper justification for that amount. Yes, the application fee could act as a filter, but a nominal amount would do the job. While a blanket waiver might not be feasible, universities should extend the eligibility of fee waiver applications to more countries.
% \end{commentbox}

\chapter{Your Application}\label{sec:application}

\epigraph{Son, if you really want something in this life, you have to work for it. Now quiet! They’re about to announce the lottery numbers.}{\textsc{The Simpsons}}

The primary focus of the admissions committee is to \textbf{evaluate your background and interest in research} (a PhD in Computer Science is a research degree!), and determine if you would \textbf{fit into} the program. To evaluate your profile, we consider
the following key indicators.

\section{Research Ability}\label{sec:research-ability}

The most concrete evidence of research ability is having \textbf{published papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the applicant has successfully involved in research.

\begin{commentbox}[Vu:]
  Many international students mention Scopus Q1, which consists of various journals from IEEE, Elsevier, and many other publishers.  I don't know/recognize many of journals listed in Scopus Q1. This might be something to be mindful of, as \textbf{CS} faculty might not be too familiar with Scopus or journals listed there, so devote some part in your statement to discuss the significance of your papers.
\end{commentbox}

However, it is understandable that many students do not have the opportunities to publish in top places. Thus, general writings, even those under submissions or rejected, would still be good (and much better than nothing).  But be sure to upload your papers with your application and talk about them in your statement (see \autoref{sec:research-statement}).  Note that local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty. However, if you have published in such places, you should still upload them, mention them in your statement, and explain why they are good.

% In CS, conferences are important and top conferences are prestigious. You can find top CS conferences at places such as \href{https://csrankings.org}{CSRankings}, which ranks CS programs based on how their faculty publish at top \emph{conferences} (see more in \autoref{sec:ranking}).

\begin{commentbox}[Craig:]
  GMU and many other universities allow you to upload your published papers and other writing samples. In many cases, even if the papers were not published at top places, we can still determine their quality by simply skimming over the paper.
\end{commentbox}

Additionally, \textbf{work experiences at renowned research laboratories}, such as Microsoft Research, can strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you should explicit say something about them in your statement.

\begin{commentbox}[Hung:]
  The reputation of VinAI has been increasing steadily over the past few years; many of my colleagues heard about VinAI.
\end{commentbox}

Finally, \textbf{participating internationally recognized competitions} can also demonstrate your research potential.
For example, participating in Math Olympiads if you want to do theory or  winning ACM programming contests if you want to ``build'' systems, e.g., software analysis.

\begin{commentbox}[Thanh:]
  Due to academic culture, professors in Vietnam usually aim for (international) journals instead of conferences. Could you give some tips on how to know whether a journal is good (CSrankings, unfortunately, only consider conferences)?.
  \tcblower
  Vu: One way is looking at what well-known researchers publish at. For example, if you are interested in a field X, you can use CSRankings to look at active faculty in X, and then look at their websites to see what journals they publish at.
\end{commentbox}

\section{Letter of Recommendation (LOR)}\label{sec:lor}

\begin{center}
  \includegraphics[width=0.6\linewidth]{files/c6.png}
\end{center}

Most PhD programs will require at least \textbf{two LORs}. Having a strong letter from an internationally recognized researcher will greatly strengthen your application. In fact, a strong letter from a well-known researcher can outweight the lack of publications.

However, obtaining such letters
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, postdocs who mentored you) that know you well enough to talk about \emph{your research experience and capabilities}. Many students get letters from supervisors from companies where they did internships or are working at. It is OK as long as it is a research-based personalized letter (once again, we are talking about PhD applications, not MS).

\begin{commentbox}[Vu:]
  If your grading system is not US standard or you are from a good school unknown outside of your country, you can ask your reference writers to explain about that in their letters.  For example, "Bach Khoa" are the top universities in Vietnam for STEM studies but few people outside Vietnam know about them.  So if you are from there, you should ask your reference writers to mention that.
\end{commentbox}

\paragraph{Self-written Letters} Many international students write letters themselves, typically due to the requests of their profs., and have them signed by their profs. Such letters have little values and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?.  Instead of the reference writer talking about you, in this case it is you who write about yourself (and they just sign the letter).

\paragraph{Generic Letters} When the writers do not know much about the applicants (e.g., just taking some course with them or not making any impression to write about), they might write a generic letter, which is not useful and also considered weak.  It might be a good idea to directly ask if the prof. is willing to write a \emph{strong} letter for you. If not, then you should ask someone else.  For example, if a student I don't know well ask me to write letter for them, I will explicitly tell them I don't know them that well to write much about them, and such a short, generic, and weak letter will not help their case.

\begin{commentbox}[Hung:]
  A sad reality is that most professors in Vietnam \textbf{DO NOT} know how to write a good letter, or are lazy in writing letters hence delegate the writing to the students. Unfortunately, there is no easy solution to this problem.
\end{commentbox}

\begin{commentbox}[Vu:]

  Another sad reality is that some professors are unwilling to write letters or write weak ones because they do not want (good) students to go abroad or only go to places where they want the students to go to. If you are in this situation, you should find someone else to write for you.
  \tcblower

  Sometimes students would go through great length just to get letters from well-known professors in their school (e.g., department head). But as mentioned, if these professors do not know you, the letters are generic and carry little value, in fact, \red{red flags}. Moreover, a top professor in Vietnam might not be well-known to US faculty (see more details in \autoref{sec:your-school}). So save the trouble and just get letters from \emph{any} professors/supervisors who knows you well and can write a good letter about \emph{your} research ability. It's better to have a good personalized
  letter about your own research ability from someone who is less
  well-known than a generic/weak letter from a well-known person.

\end{commentbox}

\paragraph{Waiving Your Right}  Choosing not to look at a reference letter is pretty standard in school and job applications. When you waive your right to see the letter, it adds a layer of trust, showing you're confident in your choice of referees and that you're not trying to twist their words. It's also about keeping things open and honest between you and your letter writers, and encourages them to be real about your strengths and qualifications. Plus, it keeps things private.

Reviewers might raise concerns (a red flag) about  a letter that is not waived, e.g., if you do not trust your letter writers, then you should find someone else to write for you. In short, it's a standard practice and a way of keeping things straightforward and respectful in the whole recommendation game.

\begin{commentbox}[Didier]
  \emph{Should letter writers have PhDs?}  In Rwanda, a lot of students interact more with teaching faculty who might not have PhD.
  \tcblower
  \textbf{Vu}: This is an interesting and useful detail that US faculty might not be aware of and students should mention about this in their statements. In general, I think it should be fine as long as that person can properly evaluate your research ability.
\end{commentbox}

\paragraph{Reminding Writers} After you submit your application, you can tell your writers that and remind them to submit their letters if they haven't done so. But don't send too many reminders as that can be annoying to the writers.

Note that most places only have deadlines for the applicant, but are very flexible with the letter writers (in many cases do not even give them any deadline).  Also, many places do not begin the admission review process right after the deadline and work on application reviews in the next semester (mid January).

\paragraph{VEF 2.0} For Vietnamese students, it's worth mentioning about the \href{https://vef2.org/}{VEF2.0 program}, which has helped many good students in gaining admission to top CS PhD programs. VEF2.0 follows an interesting model where US faculty members from leading institutions are invited to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily confined to the interview itself), these reference letters are generally effective as they have helped many students getting into top PhD programs in the US.

\section{Statement of Purpose (SOP)}\label{sec:research-statement}

\begin{center}
  \includegraphics[scale=0.4]{files/c2.png}
\end{center}

While you might not have control over LORs or \hyperref[sec:your-school]{where your go to school}, you do over your
statement of purpose (SOP) or personal statement\footnote{Some schools separate these documents and ask you to write both: SOP, which focuses on research experiences, and Personal Statement, which is everything more personal, e.g., why PhD, challenges, diversity, etc}! So write it well because it could make a big difference.
In your statement, you have the opportunity to make your application stand out and unique, which can make you \emph{fit} the CS program you're applying to, even if you don't have very strong research experience.
A well-written SOP also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \autoref{sec:funding}).
% is important because if you need (GTA) funding, it will provide evidence
% that you can teach and communicate with students.

There are various guides on writing statement, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements} are available. So I will not talk too much about statements. In short, focus on your research goal and vision, and convince us that you can achieve it through your experience, e.g., published papers, or if you work on some projects by yourself, talk about it. Also, talk about things that adcom members might not know about, e.g., your Github project with 1K+ stars or your regular contributions to well-known open-source projects. Finally, talk about what makes you different than the other 1000 applicants that we are considering (e.g., see \autoref{sec:improve-your-chance} for increasing your admission chance).

\begin{commentbox}[Hung:]
  I think asking for a “research vision” from a Ph.D. applicant is too much. Even people graduated with a Ph.D. have a great difficulty in explicating their research vision. In my opinion, you should focus on showing (not telling) your research passion (if you do have).
\end{commentbox}

Finally, this is something easy to do, but is neglected by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply \emph{here}? provide names of professors who you're interested in (if they are not already in the adcom, your application might get forwarded to them for evaluation; and they will likely be those who interview you).
This shows that you're serious and have done homework on places you're applying to.
Adcom will look for this part at the end of the statement, so do not skip it.

\begin{commentbox}[Vu:]
  I often read the research statement (and the LORs) first. If I am
  persuaded by then, I would skim over other factors and advocate for
  admission (unless I see red flags in other parts). If I am not
  convinced, then I will likely recommend rejection (unless I see
  something standout in other parts).
  \tcblower
  Do careful research on professors, don't mention \emph{emeritus} or  adjunct faculty (see more about various types of faculty in \autoref{sec:faculty-types}).
  Also, be careful not to send statements to wrong schools or mixing
  facts (e.g., talking about school X but mentioned about working with
  profs. at school Y; and definitely do not talk about George Washington when applying to George Mason). I have seen such statements more time that I
  should.
\end{commentbox}

\subsection*{Additional Resources}
\begin{itemize}
  \item \href{http://www.pl-enthusiast.net/2022/10/03/how-to-write-a-grad-school-personal-statement/}{How to Write a Grad School Personal Statement} by Mike Hicks
  \item     \href{https://cs-sop.notion.site/cs-sop/CS-PhD-Statements-of-Purpose-df39955313834889b7ac5411c37b958d?p=f5d5980a71524ebaa4e6ae57266b847c&pm=s}{CS PhD SOP database} by cs-sop.org
\end{itemize}
% \subsection{Outline of a LOR}\label{sec:lor-outline}

\section{Your School}\label{sec:your-school}

Graduating from top universities \emph{that we recognize} helps. For example, if your school is well-known, then it is \emph{``top foreign''}, which is definitely a plus.
However, if we do not know much about schools in your country, then we are uncertain about the quality of your school and  likely treat your school as \emph{``unknown foreign''}, which can be a minus point.

So what can you do about this? several things including asking your CS dept to put itself on CSRankings (it's the easiest way to get CS people to know about the school and its CS faculty)  and explaining about your school in your statement (ask your LOR writer to do that too). Of course, if you're interested in working with Vietnamese, consider  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{CS PhD programs in the US that have Vietnamese professors}.

\begin{commentbox}[Vu:]
  Sometime PhD adcom in the US will share a document such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Foreign-Top-Schools}{this one}, which lists the top schools in several countries. In some cases we ask other faculty and students if we think they know about the place.  For example, when I was a postdoc at UMD, members in their CS PhD adcom ask me to evaluate applicants from Vietnam.  During my time at UNL and now here at GMU, I have looked at Vietnamese applications (whether they are assigned to me or not) and provide input to their reviewers, e.g., X is the top tech school in Vietnam and so it should be \emph{top} instead of \emph{unknown} foreign, which makes a huge difference.
\end{commentbox}
\begin{commentbox}[Deepak:] 
  If an applicant is anxious about their school not being known outside their country, they can provide information about their school and department, with independent sources where such information can be verified.
\end{commentbox}

\section{Grades and GREs}\label{sec:grades}

Having good grades is good, but unless your school is well-known, having top grades or rankings
usually will not help because we cannot evaluate them.

This can be an issue for students in many top international universities where the competition is so high that very good students can still have low rankings from these schools (and be overlooked by the adcom).
So what to do with this? as mentioned in \autoref{sec:your-school} you can put a note about this in your statement and ask your LoR writers to talk about it.

Note that while having good grades at unknown school might not help,
having bad grades will be \red{red flag} (unless your LORs or
statement give a proper explanation). This is especially true if you
have bad grades in relevant courses (e.g., Math and CS).

\begin{commentbox}[Thanh:]
  Vietnamese universities typically offer specialized programs, such as the talented engineer program at HUST, that have highly competitive entrance exams and a limited number of available slots (e.g., 30 per year). However, these programs often set higher requirements for students, including more demanding tests and assignments, resulting in lower GPAs and overall rankings. For example, an 3.5 GPA students from such talented programs are typically much better than a 4.0 GPA students not in those programs.  Similarly, variations in GPA standards exist among different universities, with technical universities generally having lower GPAs than economical universities. These make gaining admission in the US difficult as US faculty are not familiar with these issues.
  \tcblower
  \textbf{Vu}: Vietnamese students and even faculty often lament how this grading system hurts Vietnamese students applying abroad. One way to mitigate this is making these issues known in your SOP.  Universities with Vietnamese profs are probably aware of them, but in general your letter writers \emph{and} you can explicitly mention these in their letters and your statement.
\end{commentbox}

\paragraph{GRE} Most good CS PhD programs in the US \textbf{no longer require GREs}, so you don't need to take them. However, they might be useful for international students from programs we are not familiar with.  If you have good GRE scores then you should include them in your application.

\paragraph{English Test} Unless your degrees are from specific countries such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{these}, you will need to
take standardized English test. Just do well enough to pass minimum requirement set by the university, which has many options for you to select from.

\begin{commentbox}[Vu:]
  The minimum for GMU (being above this might not mean much, but below is a \red{red flag}).
  \begin{itemize}
    \item GPA: $\ge 3.0$ in your undergrad (but we also consider the rank/prestige of your school)
    \item GRE: not required, though it can help boost your profile
    % \item but if you want to use it, then we expect a total (V+Q) of $\ge 311$ (with a $\ge 157$ Q) and A $\ge 3.0-3.5$.
    \item English requirement tests (one of the below)
          \begin{itemize}
            \item TOEF: 88 pts in total AND $\ge 20$ points in each subsection OR
            \item IELTS: $\ge 6.5$ OR
            \item DuoLingo Graduate English: $\ge 120$ OR
            \item Pearson Test of Academic English: $\ge 67$
          \end{itemize}
  \end{itemize}
\end{commentbox}

\section{CV/Resume}
The CV/Resume should provide a summary of the applicant's achievements.  It should allow the reviewers to quickly scan to identify major achievements (e.g., Publications, Programming Competition Awards, Teaching Experience).

\section{Interview and the Waiting Game}\label{sec:interview}

After you submitted your applications, the waiting game begins! For many students, this is an excruciating time. This section provides some information and tips to help you get through this time.

% but you should try to relax and focus on your study and work.  You can also use this time to improve your profile, e.g., by working on research projects, publishing papers, or improving your connections for better LoR writers.

\paragraph{Interviews} Just like a job application, you can get interviews from some schools. The most common case is that a professor is interested in working with you and want to chat to make decision (e.g., to offer RA).
The professor might include one of their current students as part of the interview.

In some case, the interview is done by several professors, e.g., to see if a student fits to their group or to recruit a very strong student to their program. The interview can also be done by adcom members to see if the student fit their program.

Typically, an interview takes about 15--30 minutes, and one important aspect of evaluation is your ability to effectively communicate, including speaking and understanding English. A professor might want to know more about your research background or even ask you to read a paper (so be sure to do it!). During the interview, you will have the opportunity to ask questions about the university. It's essential to prepare some thoughtful questions, similar to a job interview.

\begin{commentbox}[Vu:]
  At GMU, faculty are encouraged to interview candidates. For very strong candidates, the interview is actually to recruit them.  In some cases a faculty interviews a candidate that they see potentials and want to advocate for their admission. Without the interview, such applications may be more likely to be rejected.
  \tcblower
  In short, getting an interview is a good sign; it means that someone is considering you. If we are not interested in your application, we will not proceed with an interview.
\end{commentbox}

\paragraph{Not getting interviews} While it is generally good to get interview, do not feel so despair if you do not get any.  Many programs do not have the tradition of interviewing applicants. For example, at GMU, most of our admitted students with TA do not go through any interviews.

\paragraph{Timeline} The timeline for interviews and notification letters varies.  Faculty setup interviews based on their (very busy) schedule. So do not be surprised if you get an interview invitation in the last minute (also don't get surprised if your interview is on a Uber ride!). And again, some schools do not do interviews at all.

Some schools send out admission letters in batches, some do rolling admission, and some do not send anything out (e.g., you're on their wait list).  However, in general, you should hear back from most schools by around March.

In short, you should not worry too much about the timeline. Also do not send emails asking about interviews or status.

\paragraph{Updating your profile} In general, you should not send emails to update your profile.  However, if you have new publications or other achievements, you can send an email asking to update your profile.

\section{Preparing and Tracking Applications}

\paragraph{Shared Spreadsheet} Many students use a (shared) spreadsheet, e.g., Google Sheets/Docs, to help them and their letter writers to keep track of their applications. Here are some information to put on the sheet.
\begin{itemize}
  \item \textbf{Your info}: Full name, email, phone, link to website/CV.  This is helpful for the writers just in case they want to quickly get some information about you.

  \item \textbf{Applications details}: University, Dept, Application System URL, Submission Deadlines, Application Status (e.g., submitted,  rejected, wait-listed, accepted)

  \item \textbf{LoRs}: Writer 1, Writer 2, Writer 3.  If this is a shared document, you might want to omit names and just use writer X. Under each is their status, e.g., sent/not yet/reminder needed, etc.

\end{itemize}

You can then share this sheet with your reference writers and remind them to submit LoRs on your behalf (see \autoref{sec:lor}). Also, you should periodically update your writers with your status; a simple note such as \emph{``I am getting interviewed by prof. X at school Y. Do you have any advice?''} can get you a lot of help.

% \paragraph{Communicating with LoR writers and People Who Support You}

% stressfull time, rely on support from others
% ask ppl to review your SOPs
% provide information for LORs, ..
%

\chapter{Getting Admitted}\label{sec:accepted}

\epigraph{"Oh... and how is education supposed to make me feel smarter? Besides, every time I learn something new, it pushes some old stuff out of my brain. Remember when I took that home wine-making course and I forgot how to drive?"}{\textsc{The Simpsons}}

By around March you should hear back from most PhD programs you applied to. If you haven't heard back, reach out via email and ask about the status of your application.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider include the reputation of schools and professors (\autoref{sec:schoolsandprofs}), and funding availability (\autoref{sec:funding}). You will have to make your decision by around \emph{April 15}.

\paragraph{Open House} Most schools will have \textbf{Open House} events, which are a great resource to learn about the school, department, faculty, research, living, etc. Even if you can't come in person, you should attend virtually and meet with individual faculty.
During the Open House, you get a chance to talk to individual faculty and current students.  Take notes of faculty who make you excited, count those that are taking in new students (if they meet you, likely they are considering new students!).  Talk to students about their advisers, the dept, the area, funding situation etc.  Ask about anything you want to determine that they deserve \emph{you}.

\begin{commentbox}[Vu:]
  GMU has \emph{Virtual} Open House (VOH), e.g., \url{https://cs-gmu.github.io/cs-phd-voh-s23/}. We invite all admitted PhD students to the VOH through Zoom to learn about the CS program, the department, GMU, and the DC area in general. Students also get opportunities to chat with professors and current students.
\end{commentbox}

\paragraph{If you do not get admitted} If you do not get admitted to any schools or don't want to go to the ones that admit you, try again next time.  Graduate admission can involve randomness and noise.  Don't bother asking for feedback, you will not likely get any.  In the meantime, you can work on improving your profile, e.g., get more research experiences, publish more papers, improve your connections for better LoR writers, etc.

You can also consider applying to MS programs, which are typically easier to get in (but you need to pay).  If you get in an MS program at a school of your choice, you can contact professors to work with them. If you do well, you can ask the professor to support you to convert to PhD.  This is a common path for many students.

\chapter{Funding}\label{sec:funding}

\epigraph{Bart, with \$10,000, we’d be millionaires! We could buy all kinds of useful things like…love!’}{\textsc{The Simpsons}}

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!  
In the US, the common types of funding for PhD are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is an independent funding that can come from a school, a company, or an organization. \autoref{tab:funding} summarizes the differences.
Note that funding is typically more available for PhD students than 
Masters.

\begin{table}
  \centering
  \footnotesize
  \caption{Different types of PhD funding. All Covered: includes tuition, insurance, and stipend.}\label{tab:funding}
  \begin{tabular}{c|c|c|c}
    \toprule
    &\textbf{TA}&\textbf{RA}&\textbf{Fellowship}\\
    \midrule
    \textbf{From} & School & Profs. & School/External\\
    \textbf{For}                  & Teaching Assist.       & Research                        & Research                              \\
    \textbf{Cover All?} & Yes                      & Yes                             & Yes                                   \\
    \textbf{Summer?}              & No                       & Maybe                           & Yes                                   \\
    \midrule
    \textbf{Pros}                 & Research Freedom         & Get to do research              & Research Freedom                      \\
    \textbf{Cons}                 & Teaching Duties           & Research Restriction & Competitive, Limited             \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Graduate Assistantship (TA/RA)}
The most common type of funding is \textbf{graduate assistantship}, which is either TA or RA. Both TA and RA come with tuition waiving (you don't have to pay tuition), health insurance (this takes care of your insurance, which is a must have in the US), and most importantly, your stipend (i.e., your salary). Some universities also give significant discount or pay insurance for spouse/children.

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on various factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer through funding from your professor or fellowship (typicaly no TA over the summer). Fourth, like most source of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipend, but they could have higher "activity" or some other hidden fee, or you may be required to pay fees for each credit hour.

Students often complain their stipend being too low, but it is actually not bad and you can live comfortably with it.  In many cases, it is also enough to support your spouse and kids (many CS PhD students have their family with them). So don't worry too much about stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at GMU, every year we discuss about improving the benefits, and especially stipend, for our graduate students.

For a full breakdown of how much a graduate student costs, see \autoref{sec:ra-cost}.

\begin{commentbox}[Vu:]
  TA and RA at GMU have similar benefits in tuition waiving and insurance.  The college and department will set the rate for 9-month graduate assistant stipend.  TA, which is paid by the department, will likely be that amount but RA might be higher depending on the stage of the student (1st year vs ABD\footnote{All but dissertation: really close to graduate.}) and the prof.
  \tcblower
  Having health insurance is required at many US universities.  Do not assume that you're young and healthy and ignore insurance.  At GMU, and at most good CS PhD programs, your GTA or GRA \emph{will always} come with full insurance. In fact, at GMU your spouse/children will get significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

\subsection{Teaching Assistant (TA)}\label{sec:ta}

TA is common in the beginning when you haven't found an adviser who would pay you RA. As a TA, you spend up to 20 hrs/week and help with classes (e.g., grading or teaching labs/recitation). Your TA is paid through the school or department, i.e., they hire you to help teach.  During a semester, a TA might work with several courses and professors (not necessary their adviser).  TA funding is not typically available during the summer, which has few courses.

\paragraph{How to get TA?}  Unless you have other funding such as RA or Fellowships, TA is typically default for good CS PhD programs. When you apply to be a full-time student,  state that you need financial assistant. It is common that the PhD committee will either admit you and give you GTA, or reject you; i.e., we do not admit a student without supporting them.

\begin{commentbox}[Vu:]
  At GMU CS, students admitted with TA have  4 years of GTA guaranteed, and in some cases also receive  stipend for the first summer.
\end{commentbox}

Even if you have other funding and do not need TA, you still should do TA at least once.  This allows you to see what teaching is like, which is especially helpful for research career where you often give talks and tell people about your work. GMU sometimes has classes that a more senior student can teach.  In that case, you will be paid as GTA or even sometimes as a lecturer.  This is a good opportunity for students to get teaching experience and also get paid more (as a lecturer).

\subsection{Research Assistant (RA)}
RA is provided through a professor through their own funding so you can work on their project.  
You do not need to teach as an RA, so you can focus on your research. Depending on the professor, RA may be available during the summer. \autoref{sec:ra-cost} gives more details on RA budget.

\textbf{How to get RA?} When a professor recruits you, they will likely give you RA right away (e.g., when you apply).  A common scenario is that you first get admitted with TA, and then after a year or two find an adviser to support you with RA. See how to contact a prof. in \autoref{sec:contact} for research opportunities.

\begin{commentbox}[Vu:]
  If you got recruited by a prof., who would give you RA right away, it's very likely you will get admitted.  For example, if a prof., even if not in PhD adcom, wants to work with and funds you, the PhD adcom will respect that decision and admit you (unless your application has many red flags).
\end{commentbox}

\section{Fellowship/Scholarship}

Fellowship is another type of funding that students can get (e.g., from the university, industries, government).
Fellowships are typically competitive and generous, and gives pretty much all benefits tuition/insurance that a TA/RA has.  Moreover, they often give higher stipend (including summer) and open doors for job opportunities (e.g., internship).

\textbf{How to get Fellowship?} You need to apply for them.  Many schools provide such fellowships to attract admitted students.
External fellowships including from the US government and tech companies (e.g., Google, Microsoft, Facebook).  Fellowships from US government would likely require US citizenship or residency. % However, tech companies including Google, Microsoft, Facebook have fellowships that international students can apply for.

Prestigious external fellowships typically require a clear and good research plan, so it is a good idea to wait until at least your second year to have research experience and even publication before applying. Remember, you're competing with the top PhD students at top universities worldwide.

\begin{commentbox}[Vu:]
  PhD applicants at GMU are automatically eligible for a \emph{Presidential Fellowship}.  It is at least as good as GTA but the most important thing is that as a fellowship it is truly free money (i.e., you are not depending on any prof. or TA).  PhD adcom members nominate applicants for this fellowship and the whole committee will decide.
\end{commentbox}

In general, fellowship, especially competitive ones from the government or company, is prestigious, and you will stand out if you get one.  Every PhD student has pubs, but only superstars have NSF grad or Microsoft fellowship. In fact, these are so prestigious that even if you didn't get it but make it to the final round or even ``honorable mentioning'', school will still mention you on their website and you still should put it on your CV.

\chapter{Choosing Schools and Professors}\label{sec:schoolsandprofs}

Choosing a school and an adviser is clearly among the most important things in your mind when you apply and especially when you get admitted.  This is further complicated due to cultural differences and unfamiliarity of  international students to the US higher education system.  This section aims to mitigate some confusions and help you make informed decision.

\section{Choosing a University}

We will first discuss about universities in the US that offer PhD in CS. Then we will talk about how to rank and select them.

\subsection{Schools that offer PhD in CS}

Most universities the US have CS programs. 
However, while many universities offer PhD in their CS programs, many do not.  These universities might offer just Bachelor degrees (e.g., BS) and no graduate studies (i.e., no MS or Ph.D degrees), or they just offer MS programs (but no PhD). For example, Penn State in University Park has Ph.D. in CS,  but Penn State in Harrisburg only has BS and MS in CS, and Penn State in York only has BS in CS.  On the other hand, multiple locations of the University of Texas, e.g., Austin, Dallas, Arlington, have PhDs in CS.

Thus, if your goal is Ph.D. in CS, you have to aim only for schools offering such a degree.  %This might sound obvious but many students (even those in the US) get confused due to the large number of universities. 
While this can be confusing due to the large number of universities in the US, a little research, e.g., searching for PhD in CS from the school website, will help you find out. All schools listed in \autoref{sec:ranking} have PhD studies in CS, so you can start there.

% \subsection{R1, R2, ...}

>>>>

\mainmatter

\chapter{Research}
\chapter{Obtaining Funding}
\section{NSF and other Government Funding}

\paragraph{Other Government Funding}

\section{Internal Funding}

\section{Industrial Gifts}

\chapter{Teaching}
\section{Teaching Load}
\section{Teaching Evaluation}
\section{Teaching Philosophy}
\section{Teaching Awards}

\chapter{Mentoring}
\section{Mentoring Undergraduate Students}
\section{Mentoring Graduate Students}
\section{Mentoring Postdocs}
\section{Mentoring Junior Faculty}

\chapter{Services}

\section{Service to your Professional Community}
\section{Service to the Department}
\section{Service to the University}
\section{Service to the Community}

\section{Serving on NSF Panels}
You should serve on NSF panels.  It is a great way to learn about the grant process: how to distinguish between good and bad proposals and how people review propsoals.
It is also a good way learn what other people are doing, what is hot, and what is not.
